Formalizing Stack Safety as a Security Property

### 2023-01-29 submission5/ideas-v2.tex
Nits

\subsubsection{Deferred and Lazy Properties}
%
We seek to support both \emph{eager} property enforcement (where potential
violations are detected immediately) as well as more permissive \emph{lazy}
properties (as motivated in \cref{XXX}).
%
Micro-policies are well suited to both types of properties. There is an
interesting connection between lazy enforcement and the deferred checks of stack
canaries and shadow stacks. Other techniques, like split stacks and bounds
checking---and the closely related hardware capabilities---are fundamentally
eager techniques.

\subsubsection{Low-Level Component Support}
%
Our framework considers a \emph{single-, low-level view} of all program code and
data; a key advantage of this approach is that it allows us to remain
\emph{language-agnostic}.
%
Some software-based protection mechanisms are designed specifically with
a \emph{high-level language} in mind, e.g. bounds checking and split stacks.
They tend to offer stronger security guarantees than other software techniques.
In exchange, their threat model becomes weaker, as they make additional
assumptions about the admissible behavior of program components
%
(for example, bounds checking only applies to code that has been instrumented by
a \emph{secure compiler} \ref{}).
%
In general, a multi-level extension of the framework offering adequate reasoning
capabilities for this class of policies would need to model multiple source
language and a notion of code provenance---a reflection of the call stack as the
melting pot of language abstractions.
%
The common case of one source and one target language mediated by a compiler may
be solved by a simpler lifting of our framework from the machine to the source
level \rb{e.g., SoftBound could use a C language semantics and Tagged C for
security modeling (lifted properties would likely hold under the expected
assumptions)}.

@@

\subsubsection{Software vs. Hardware}
%
One first general observation about software-based techniques is that they
are \emph{less modular} than their hardware counterparts, that is, the
countermeasures are more tightly coupled with the program code they protect.
%
By contrast, although programs must effectively leverage the protection
mechanisms offered by the hardware, the distinction between those protections
and the program logic itself is much cleaner.
%
Our framework relies on a \emph{separation} of security-relevant operations and
regular program code \rb{stress that any scalable, composable solution must do
this!}.
%
While the various extensions suggested in this section could help abstract the
complexities of software safeguards, hardware techniques remain the most elegant
and general exemplars for the methods we seek to demonstrate.
%
\rb{(This is also related to our single-level model)}

\subsubsection{Formal Models}
%
A second general observation about software-based techniques is that they tend
to be \emph{less absolute}: they enforce weaker properties or under stronger
assumptions, and their threat models are usually fairly informal.
%
%
The closest previous attempt
to produce a formal characterization of stack safety does so using a {\em fully abstract overlay semantics},
a style of correct-by-construction machine \cite{SkorstengaardSTKJFP}.
But this approach
builds the security properties into the semantics of the machine without externalizing them or their meaning.
It is also not suitable for the {\em lazy} enforcement mechanism
of Roessler and DeHon, in which stores are not checked, and illegal stores are
only caught when their location is later read.
%
\rb{Need to discuss overlay semantics, but also note that it is very intensional
  and very specific, as the authors acknowledge. I'd introduce this a bit more
  gradually. Also, if phrased this way, it sounds like its main drawback is a
  lack of generality, to which we should contrast a more general approach, not
  just one that is specialized to a specific policy, i.e. lazy micro-policies
  (I'd try to present these more gradually as well).}
%
\rb{Also, testing certainly (and motivate), but not only?}

Overall, the generality and flexibility of micro-policies, arising from an
elegant design that clearly separates the program from its security, are the
most appropriate choice for an exploration of this new framework.
%
\rb{FIXME: Finish motivation. We focus on micro-policies, but\ldots Something
  about existing policies, exercising the framework most generally,
  representative experiments?}

@@

A {\it security semantics} extends a machine
with additional context about the identity of current and pending
functions (which act as security principals) and about the registers and memory they require
to be secure. This added context is purely notional;
it does not affect the real machine. The security context
evolves dynamically through the execution of {\it security-relevant operations},
which include events like calls, returns, and frame manipulations.
Our security properties are phrased in terms of this context, often as predicates
on future states, e.g. of the form ``when control returns to the current function...'',
or relations on traces of future execution (hyper-properties).

@@

Questionably that they are strictly more realistic in the sense that it is very
costly or requires special sophisticated/experimental types of capability not
available/perhaps not very feasible to implement.

@@

Problems of adapting this work to capability machines: security semantics tightly

>>>>

\subsubsection{Deferred and Lazy Properties}
%
We seek to support both \emph{eager} property enforcement (where potential
violations are detected immediately) as well as more permissive \emph{lazy}
properties (as motivated in \cref{XXX}).
%
\rb{(Mention complexity/relation between eager/lazy properties somewhere? Later?)}
%
Micro-policies are well suited to both types of properties. There is an
interesting connection between lazy enforcement and the deferred checks of stack
canaries and shadow stacks. Other techniques, like split stacks and bounds
checking---and the closely related hardware capabilities---are fundamentally
eager techniques.

\subsubsection{Low-Level Component Support}
%
Our framework considers a \emph{single-, low-level view} of all program code and
data; a key advantage of this approach is that it allows us to remain
\emph{language-agnostic}.
%
Some software-based protection mechanisms are designed specifically with
a \emph{high-level language} in mind, e.g. bounds checking and split stacks.
They tend to offer stronger security guarantees than other software techniques.
In exchange, their threat model becomes weaker, as they make additional
assumptions about the admissible behavior of program components
%
(for example, bounds checking only applies to code that has been instrumented by
a \emph{secure compiler} \ref{}).
%
In general, a multi-level extension of the framework offering adequate reasoning
capabilities for this class of policies would need to model multiple source
languages and a notion of code provenance---a reflection of the call stack as the
melting pot of language abstractions.
%
The common case of one source and one target language mediated by a compiler may
be solved by a simpler lifting of our framework from the machine to the source
level \rb{e.g., SoftBound could use a C language semantics and Tagged C for
security modeling (lifted properties would likely hold under the expected
assumptions)}.

@@

\subsubsection{Software vs. Hardware}
%
One first general observation about software-based techniques is that they
are \emph{less modular} than their hardware counterparts, that is, the
countermeasures are more tightly coupled with the program code they protect.
%
By contrast, although programs must effectively leverage the protection
mechanisms offered by a hardware-based mechanism, the distinction between those protections
and the program logic itself is much cleaner.
%
Our framework relies on a \emph{separation} of security-relevant operations and
regular program code \rb{stress that any scalable, composable solution must do
this!, add fwd ref?}.
%
While the various extensions suggested in this section could help abstract the
complexities of software safeguards, hardware techniques remain the most elegant
and general exemplars for the methods we seek to demonstrate.
%
\rb{(This is also related to our single-level model)}

\subsubsection{Formal Models}
%
A second general observation about software-based techniques is that they tend
to be \emph{less absolute}: they enforce weaker properties or rely on stronger
assumptions, and their threat models are usually fairly informal.
%
The closest attempt
to produce a formal characterization of stack safety does so using a {\em fully abstract overlay semantics},
a style of correct-by-construction machine \cite{SkorstengaardSTKJFP}.
But this approach
builds the security properties into the semantics of the machine without externalizing them or their meaning.
It is also not suitable for the {\em lazy} enforcement mechanism
of Roessler and DeHon, in which stores are not checked, and illegal stores are
only caught when their location is later read.
%
\rb{Move the bit about laziness to the corresponding point above? Also emphasize
  (and contributions?) that we go further by having an extensional, executable
  specification, which we can test or verify.}
%
\rb{Need to discuss overlay semantics, but also note that it is very intensional
  and very specific, as the authors acknowledge. I'd introduce this a bit more
  gradually. Also, if phrased this way, it sounds like its main drawback is a
  lack of generality, to which we should contrast a more general approach, not
  just one that is specialized to a specific policy, i.e. lazy micro-policies
  (I'd try to present these more gradually as well).}
%
\rb{Also, testing certainly (and motivate), but not only?}

Overall, the generality and flexibility of micro-policies, arising from an
elegant design that clearly separates the program from its security, are the
most appropriate choice for an exploration of this new framework.
%
\rb{FIXME: Finish motivation. We focus on micro-policies, but\ldots Something
  about existing policies, exercising the framework most generally,
  representative experiments? This kind of repeats the opening of the section.}

@@

A {\em security semantics} extends a machine
with additional context about the identity of current and pending
functions (which act as security principals) and about the registers and memory they require
to be secure. This added context is purely notional;
it does not affect the real machine. The security context
evolves dynamically through the execution of {\em security-relevant operations},
which include events like calls, returns, and frame manipulations.
Our security properties are phrased in terms of this context, often as predicates
on future states, e.g. of the form ``when control returns to the current function...'',
or relations on traces of future execution (hyper-properties).

@@

@@

One may wonder whether capability methods are strictly more realistic in the sense they are very costly
costly or require sophisticated/experimental types of capability not
available/perhaps not very feasible to implement---it's an open question.
Challenges of adapting this work to capability machines: security semantics tightly

### 2023-01-29 submission5/intro-v2.tex
Nits

Computer programming at any scale relies on \emph{functions} (variously known as
subroutines, procedures, methods, etc.) as a \emph{foundational abstraction} to
structure programs into building blocks representing units of computation that
can \emph{call} one another to create larger computations.
%
At a low level, a function call comprises its own data (private or shared), as
well as information about its \emph{caller} and \emph{callee} functions. The
runtime must keep track of the local contexts of all active function activations
at any given time; the \emph{fundamental data structure} charged with realizing
this abstraction is the \emph{(call) stack}.
%
\rb{This should be well known, so no need for illustration?}

The correct execution on a program---and its security---is contingent on all of
its functions properly implementing this abstraction by means of a careful use
of the call stack. Be it accidental or intentional, incorrect management of the
stack is at the root of a myriad of faults whose effects range from (rarely)
innocuous to (usually) catastrophic.
%
From a security perspective, attacks on the stack (say, from a malicious
function that chooses to abuse its access to the contents of this data
structure) are as fundamental, varied and pervasive as the functional
abstraction itself.
%
\rb{Begins to prefigure the idea of multiple stack implementations.}

@@

The protections offered by stack protection mechanisms are commonly described by giving
concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more formal characterization to the reader's intuition.
%
To make these notions more precise, we review a selection of specimens from the
menagerie and begin to consider the difficulties involved in specifying and
verifying the protection they offer.
%
For an assessment of these methods in the context of our stack security
framework, see \cref{XXX} below.

@@

\paragraph{Split Stacks}
%
Code-pointer integrity seeks to prevent control-flow hijacking attacks by
splitting the program memory, including the stack, into a regular region and a
safe region; those objects that require protection are located in the latter and
all accesses protected by static or dynamic checks.
%
Like shadow stacks, bounds checking is a high-level technique designed to
protect code written in a high-level language, but, unlike them, it is meant to
offer protection under a stronger threat model that also considers the presence
of untrusted code. Similar to most software-based methods, it purports to
protect exclusively the control-relevant parts of the stack, making no claims
about its data-relevant parts.

@@

In addition to many the software-based techniques developed to provide greater
degrees of protection to the stack, often at a significant cost in performance,
some proposals have turned to specialized hardware features to offset those
penalties and attempt to offer greater levels of protection.
%
\rb{Something about how realistic/practical these alternatives are? Are we
  selecting partly based on prototypes, etc.?}

\paragraph{Capabilities}
%
\rb{(There is plenty of variety here, should we focus on the Cerise-and-friends
  (CHERI-adjacent) work exclusively?)}
%
Capability machines can represent and manipulate their homonymous
capabilities---unforgeable tokens of authority over specific regions of
memory---using a combination of dedicated out-of-band memory and a collection of
specialized ISA instructions.
%
The most prominent contemporary representative of this long line of work is
CHERI \cite{}, a modern architecture designed to provide efficient fine-grained
memory protection and compartmentalization.
%
Previous work has demonstrated the use of these general principles \rb{in
  simplified models} to implement \emph{secure calling conventions} for the call
stack \cite{}. Those conventions can be seen as combining the advantages and
protections of various families of software-based methods.
%
They are closely related \rb{too strong?} to this work; we discuss them in more
detail in \cref{XXX}.
%
\rb{HardBound is also related to this}

\paragraph{Micro-Policies}
%
\rb{(This partly overlaps with the technical preliminaries, try to offer a quick
  synthesis here.)}
%
Micro-policy machines are a family of tagged architectures that decorate
addressable memory and registers with arbitrary metadata tags, and extend the
processor with a programmable hardware monitor that uses custom rules on
metadata tags to implement security policies of various types.
%
This flexibility can be harnessed to implement a wide array of stack protection policies with relative ease, covering and going beyond the capabilities \rb{word?} of the various software-based techniques \cite{} \rb{as demonstrated by}.
%
They comprise very general forms of \emph{lazy enforcement}, which can tolerate
potential but harmless violations \rb{the system programmer's motto: ``trust me,
  I know what I'm doing''} while still detecting them if and when their effects
become tangible.
%
Micro-policies are one of the centerpieces of our technical development; we
present them in more detail in \cref{XXX}.

@@

As we begin to intuit, the security of the call stack and its contents can be
viewed through many different lenses. Each of these points of view results in a
\emph{unique specification} of the meaning of \emph{stack safety} for a
particular protection mechanism---mostly an informal specification if one is
produced, and almost always left implicit to some degree.
%
Moreover, different protections operate under different threat models, further
complicating their comparison and the assessment of the safeguards they
effectively offer their clients---the programs themselves.
%
On the face of it, it seems perplexing that there is not a \emph{standard
specification} (or set of specifications) for such a fundamental problem.
Instead of reverse engineering an specific enforcement mechanism obtain a
(necessarily approximate) specification, we turn the problem around on its head
and seek more general solutions from first principles.
%
\rb{(Problems/notes: ``backward specs'' not the most useful to their clients,
  not designed with their security needs in mind---goes in the direction of
  program logics---; to reason about security in complex systems an explicit and
  clear spec is preferable; many of the informal specs are at least
  superficially similar and overlapping; hopefully all of them can be understood
  from a small number of fundamental concepts)}

We propose a novel characterization of stack safety using the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi
%
We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety. Some focus
exclusively on control flow, some enforce integrity but not confidentiality, and many
focus on the caller over the callee.
%
\rb{Return to this after extended discussion?}
%
This gives us a total of five core properties: \(\wbcf\),
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)).
%
\rb{Forward references?}

>>>>

Computer programming at any scale relies on \emph{functions} (variously known as
subroutines, procedures, methods, etc.) as a \emph{foundational abstraction} to
structure programs into building blocks representing units of computation that
can \emph{call} one another to create larger computations.
%
At a low level, a function call comprises its own data (private or shared), as
well as information about its \emph{caller} and \emph{callee} functions. The
runtime must keep track of the local contexts of all function activations
at any given time; the \emph{fundamental data structure} charged with realizing
this abstraction is the \emph{(call) stack}.
%
\rb{This should be well known, so no need for illustration?}

The correct execution on a program---and its security!---is contingent on all of
its functions properly implementing this abstraction by means of a careful use
of the call stack. Be it accidental or intentional, incorrect management of the
stack is at the root of a myriad of faults whose effects range from (rarely)
innocuous to (usually) catastrophic.
%
From a security perspective, attacks on the stack (say, from a malicious
function that chooses to abuse its access to the contents of this data
structure) are as fundamental, varied and pervasive as the functional
abstraction itself.
%
\rb{Begins to prefigure the idea of multiple stack implementations.}

@@

The protections offered by stack protection mechanisms are commonly described by giving
concrete examples of attacks that they can prevent---corruption of return
addresses, buffer overflows, use of uninitialized variables, etc.---leaving
a more formal characterization to the reader's intuition.
%
To make these notions more precise, we review a selection of specimens from the
menagerie and begin to consider the difficulties involved in specifying and
verifying the protections they offer.
%
For an assessment of these methods in the context of our stack security
framework, see \cref{XXX}.

@@

\paragraph{Split Stacks}
%
Code-pointer integrity seeks to prevent control-flow hijacking attacks by
splitting the program memory, including the stack, into a regular region and a
safe region; objects that require protection are placed in the latter and
all accesses to them protected by static or dynamic checks.
%
Like shadow stacks, bounds checking is a high-level technique designed to
protect code written in a high-level language, but it is designed to
offer protection under a stronger threat model that also considers the presence
of untrusted code. As with most other software-based methods, it purports to
protect exclusively the control-relevant parts of the stack, making no claims
about its data-relevant parts.

@@

In addition to the many software-based techniques developed to provide greater
degrees of protection to the stack, often incurring a significant performance penalty,
some proposals have turned to specialized hardware features to offset those
costs and attempt to offer greater levels of protection.
%
\rb{Something about how realistic/practical these alternatives are? Are we
  selecting partly based on prototypes, etc.?}

\paragraph{Capabilities}
%
\rb{(There is plenty of variety here, should we focus on the Cerise-and-friends
  (CHERI-adjacent) work exclusively?)}
%
Capability machines can represent and manipulate
unforgeable tokens of authority over specific regions of
memory using a combination of dedicated out-of-band memory and
specialized ISA instructions.
%
The most prominent contemporary representative of this long line of work is
CHERI \cite{}, a modern architecture designed to provide efficient fine-grained
memory protection and compartmentalization.
%
Previous work has demonstrated the use of these general principles \rb{in
  simplified models} to implement \emph{secure calling conventions} for the call
stack \cite{}. Those conventions can be seen as combining the advantages and
protections of various families of software-based methods.
%
They are closely related \rb{too strong?} to this work; we discuss them in more
detail in \cref{XXX}.
%
\rb{HardBound is also related to this}

\paragraph{Micro-Policies}
%
\rb{(This partly overlaps with the technical preliminaries, try to offer a quick
  synthesis here.)}
%
Micro-policy machines are a family of tagged architectures that decorate
addressable memory and registers with arbitrary metadata tags, and extend the
processor with a programmable hardware monitor that uses custom rules on
metadata tags to implement many different kinds of security policies.
%
This flexibility can be harnessed to implement a wide array of stack protection policies with relative ease, covering and going beyond the capabilities \rb{word choice?} of the various software-based techniques \cite{} \rb{as demonstrated by}.
%
These policies comprise very general forms of \emph{lazy enforcement}, which can tolerate
potential but harmless violations \rb{the system programmer's motto: ``trust me,
  I know what I'm doing''} while still detecting them if and when their effects
become tangible.
%
Micro-policies are one of the centerpieces of our technical development; we
present them in more detail in \cref{XXX}.

@@

As we begin to intuit, the security of the call stack and its contents can be
viewed through many different lenses. Each of these points of view results in a
\emph{unique specification} of the meaning of \emph{stack safety} for a
particular protection mechanism---usually an informal specification if one is
produced, and almost always left implicit to some degree.
%
Moreover, different protections operate under different threat models, further
complicating their comparison and the assessment of the safeguards they
effectively offer their clients---the programs themselves.
%
On the face of it, it seems perplexing that there is not a \emph{standard
specification} (or set of specifications) for such a fundamental problem.
Instead of reverse-engineering a specific enforcement mechanism to obtain a
(necessarily approximate) specification, we turn the problem around on its head
and seek more general solutions from first principles.
%
\rb{(Problems/notes: ``backward specs'' not the most useful to their clients,
  not designed with their security needs in mind---goes in the direction of
  program logics---; to reason about security in complex systems an explicit and
  clear spec is preferable; many of the informal specs are at least
  superficially similar and overlapping; hopefully all of them can be understood
  from a small number of fundamental concepts)}

We propose a novel characterization of stack safety using the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi
%
We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety. Some focus
exclusively on control flow, some enforce integrity but not confidentiality, and many
focus on the caller over the callee.
%
\rb{Return to this after extended discussion? Forward ref?}
%
This gives us a total of five core properties: \(\wbcf\),
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)).
%
\rb{Forward references?}
